Optimal (Multiway) Spatial Joins

### 2024-06-06 msj.tex
yufei

\def\xleft{\mathrm{left}}
\def\xright{\mathrm{right}}
\def\ybot{\mathrm{bot}}
\def\ytop{\mathrm{top}}
\def\maxtop{\mathrm{maxtop}}
\def\minleft{\mathrm{minleft}}
\def\gleft{\mathrm{left\text{-}guard}}
\def\gbot{\mathrm{bot\text{-}guard}}
\def\cross{\mathrm{cross}}
\def\dcross{\mathrm{d\text{-}cross}}
\def\contained{\mathrm{contained}}
\def\cat{\mathrm{cat1}}
\def\catt{\mathrm{cat2}}
\def\out{\mathrm{OUT}}
\def\side{\mathrm{side}}

@@

Given a horizontal segment $h = [x_1, x_2] \times y$, we say that a rectangle $r$ is a {\em left-end covering rectangle} of $h$, if $r$ contains the left endpoint of $h$ (i.e., $(x_1, y) \in r$). A horizontal/vertical segment $s$ {\em crosses} a rectangle $r$, if $s \cap r \ne \emptyset$ but $r$ covers neither of the two endpoints of $s$. A rectangle $r$ {\em contains} a horizontal/vertical segment $s$, if $r$ covers both endpoints of $s$.

@@

Let $S$ be a set of segments where either all segments are horizontal or all are vertical. Given a rectangle $r$, we define 
\myeqn{
    \cross_S(r) = \{s \in S \mid \text{$s$ crosses $r$} \};
    \label{eqn:cross}
}
namely, $\cross_S(r)$ is the set of segments in $S$ crossing $r$.
Let $R$ be a set of rectangles. Given a horizontal segment $h$, we define 
\myeqn{
    \contained_R(h) = \{r \in R \mid \text{$r$ contains $h$} \};
    \label{eqn:contained}
}
namely, $\contained_R(h)$ is the set of rectangles in $R$ containing $h$.

@@

\extraspacing {\bf Problem $\bm{\mathscr{E}}$.} The input involves a set $H$ of horizontal segments and a set $R$ of rectangles. The goal is to report, for each segment $h \in H$, the set $\contained_R(h)$ --- defined in \eqref{eqn:contained} --- where the rectangles are sorted by their right boundaries. Formally, if $r_1, r_2, ..., r_z$ for some $z \ge 1$ are all the rectangles in $\contained_R(h)$, we output $(h: r_1, r_2, ..., r_z)$, provided that $\xright(r_i) \ge \xright(r_{i-1})$ for each $i \in [2, z]$. Figure~\ref{fig:probs}e gives an example where $H = \set{h_1, h_2, h_3}$ and $R = \set{r_1, r_2, r_3}$; the output is $\set{(h_1: r_2, r_1), (h_2: r_2, r_3), (h_3: \emptyset)}$. In Appendix~\ref{app:bricks}, we explain how to solve the problem in $O(n \log n + \out)$ time where $n = |H| + |R|$ and $\out$ is the number of pairs $(h, r) \in H \times R$ such that $r$ contains $h$.

@@

\extraspacing {\bf Generating $\bm{\J_2}$.} Take any $(k-1)$-tuple $\bm{t} = (r_1, ..., r_{k-3}, h', v) \in \J(R_1, ..., R_{k-3}, H', V)$. Note that $B_\bm{t}$ --- defined in \eqref{eqn:B_t} --- is the point $h' \cap v$ (the intersection of $h'$ and $v$). Suppose $h' = [x_1, x_2] \times y$ and $B_\bm{t} = (x, y)$, we define the {\em effective horizontal segment} of $\bm{t}$ as the horizontal segment $[x_1, x] \times y$. This allows us to define
\myeqn{
    \contained_{R_{k-2}}(\bm{t}) = \{r \in R_{k-2} \mid \text{$r$ contains the effective horizontal segment of $\bm{t}$}   \}
    \label{eqn:type2:contained-t}
}
The above should not be confused with \eqref{eqn:contained}, where the ``$\contained$'' function takes a segment as the parameter, rather than a tuple.

\extraspacing {\underline{\em Example 5.3.}} Consider the tuples $\bm{t}_1$, $\bm{t}_2$, and $\bm{t}_3$ of $ \J(R_1, H', V)$ given in Example 5.2. For $\bm{t}_1 = (\alpha, h_2', v_1)$, its effective horizontal segment is $\texttt{DC}$ (see Figure~\ref{fig:hv:type2:ex}b). For $\bm{t}_2 = (\alpha, h_3', v_1)$, its effective horizontal segment is \texttt{EB}. For $\bm{t}_3 = (\alpha, h_3', v_2) \in \J(R_1, H', V)$, its effective horizontal segment is \texttt{EA}. Accordingly, as can be seen from Figure~\ref{fig:hv:type2:ex}c,
$\contained_{R_{k-2}}(\bm{t}_1) = \contained_{R_2}(\bm{t}_1) = \set{\beta_1}$, $\contained_{R_2}(\bm{t}_2) = \set{\beta_1, \beta_2, \beta_3}$, and $\contained_{R_2}(\bm{t}_3) = \set{\beta_3}$.
\done

@@

\begin{lemma} \label{lmm:hv:type2:properties}
    All the following statements are true:
    \myenums{
        \item Consider any $(k-1)$-tuple $\bm{t} \in \J(R_1, ..., R_{k-3}, H', V)$. Denote by $h$ the full segment of $\bm{t}[k-2]$. Then, for any $r \in \contained_{R_{k-2}}(\bm{t})$, the $k$-tuple $(\bm{t}[1], ..., \bm{t}[k-3], r, h, \bm{t}[k-1])$ must belong to $\J_2$.

        \item Consider any $k$-tuple $(r_1, ..., r_{k-2}, h, v) \in \J_2$. Let $h'$ be the trimmed segment of $h$ and set $\bm{t} = (r_1, ..., r_{k-3}, h', v)$. Then, $\bm{t} \in \J(R_1, ..., R_{k-3}, H', V)$ and $r_{k-2} \in \contained_{R_{k-2}}(\bm{t})$.

        \item $\sum_{\bm{t}} |\contained_{R_{k-2}}(\bm{t})| \le \out$, where the summation is over all $\bm{t} \in \J(R_1, ..., R_{k-3}, H', V)$.
    }
\end{lemma}

@@

\mytab{
    \> {\bf algorithm generate-$\J_2$} \\
    \> 1.\> $\J_2 = \emptyset$ \\
    \> 2.\> {\bf for} each $(k-2)$-tuple $\bm{t} \in \J(R_1, ..., R_{k-3}, H', V)$ {\bf do} \\
    \> 3.\>\> $h \leftarrow$ the full segment of $\bm{t}[k-2]$ \\
    \> 4.\>\> {\bf for} each $r \in \contained_{R_{k-2}}(\bm{t})$ {\bf do} \\
    \> 5.\>\>\> add $(\bm{t}[1], ..., \bm{t}[k-3], r, h, \bm{t}[k-1])$ to $\J_2$
}

@@

The correctness of the algorithm follows from statements (1) and (2) of Lemma~\ref{lmm:hv:type2:properties}. Furthermore, if we are given $\contained_{R_{k-2}}(\bm{t})$ for each $\bm{t}$, statement (3) of Lemma~\ref{lmm:hv:type2:properties}
assures us that the algorithm runs in $O(1 + |\J(R_1, ..., R_{k-3}, H', V)| + k \sum_\bm{t} |\contained_{R_{k-2}}(\bm{t})|) = O(1 + k \cdot \out)$ time, where the derivation used Lemma~\ref{lmm:hv:type2:recur-output} and statement (3) of Lemma~\ref{lmm:hv:type2:properties}.

\extraspacing {\underline{\em Example 5.4.}} For $\bm{t}_1 = (\alpha, h_2', v_1)$, the full segment of $h_2'$ is $h_2$. As $\beta_1$ is the only rectangle in $\contained_{R_2}(\bm{t}_1)$, Line 5 of the algorithm adds tuple $(\alpha, \beta_1, h_2, v_1)$ to $\J_2$. For $\bm{t}_2 = (\alpha, h_3', v_1)$, the full segment of $h_3'$ is $h_3$. As $\contained_{R_2}(\bm{t}_2) = \set{\beta_1, \beta_2, \beta_3}$, Line 5 adds $(\alpha, \beta_1, h_3, v_1)$, $(\alpha, \beta_2, h_3, v_1)$, and $(\alpha, \beta_3, h_3, v_1)$ to $\J_2$. Finally, the processing of $\bm{t}_3 = (\alpha, h_3', v_2)$ adds $(\alpha, \beta_3, h_3, v_2)$ to $\J_2$.
\done

@@

We are interested in the sets $\contained_{R_{k-2}}(h^*)$ of the segments $h^*$ in $H^*$, where $\contained_{R_{k-2}}(h^*)$ --- defined in \eqref{eqn:contained} --- is the set of rectangles in $R_{k-2}$ containing $h^*$. These sets have some useful properties:

\begin{lemma} \label{lmm:hv:type2:contained-h*}
    Both statements below are true: 
    \myenums{
        \item $\sum_{h^* \in H^*} |\contained_{R_{k-2}}(h^*)| \le \out$.

        \item Consider any tuple $\bm{t} \in \J(R_1, ..., R_{k-3}, H', V)$. Set $h' = \bm{t}[k-2]$ and let $h^*$ be the minimal segment of $h'$. Then, $$\contained_{R_{k-2}}(\bm{t}) \subseteq \contained_{R_{k-2}}(h^*).$$ Furthermore, if the rectangles $r$ in $\contained_{R_{k-2}}(h^*)$ are sorted in descending order of $\xright(r)$, then $\contained_{R_{k-2}}(\bm{t})$ includes a prefix of the sorted order.
    }
\end{lemma}

@@

\extraspacing {\underline{\em Example 5.6.}} It is clear from Figure~\ref{fig:hv:type2:ex}(d) that $\contained_{R_2}(h_3^*)$ has rectangles $\beta_3, \beta_2$, $\beta_1$, sorted in descending order of their right boundaries' x-coordinates. Consider $\bm{t}_2 = (\alpha, h_3', v_1)$ and $\bm{t}_3 = (\alpha, h_3', v_2)$. Segment $h_3^*$ is the minimal segment of $h_3'$. As stated in Lemma~\ref{lmm:hv:type2:contained-h*}, both $\contained_{R_2}(\bm{t}_2) = \set{\beta_1, \beta_2, \beta_3}$ and $\contained_{R_2}(\bm{t}_3) = \set{\beta_1}$ are prefixes of the sorted order of $\contained_{R_2}(h_3^*)$. Regarding $h_2^*$, it is the minimal segment of $h_2'$, and $\contained_{R_2}(h_2^*)$ contains only  $\beta_1$. For $\bm{t}_1 = (\alpha, h_2', v_1)$, $\contained_{R_2}(\bm{t}_1) = \set{\beta_1}$ is a (trivial) prefix of $\contained_{R_2}(h_2^*)$, as is also consistent with the lemma.
\done

Finding the $\contained_{R_{k-2}}(h^*)$ sets of all $h^* \in H^*$ is an instance of the find-all-sorted version of Problem $\mathscr{E}$ (with $H^*$ and $R_{k-2}$ as the input). The cost is $O(n \log n + \out)$ according statement (1) of Lemma~\ref{lmm:hv:type1:cross-r*} and the discussion in Section~\ref{sec:bricks}. Note that, for each $\contained_{R_{k-2}}(h^*)$ computed, the rectangles $r$ therein have been sorted in descending order of $\xright(r)$.

\extraspacing {\bf Computing the ``$\bm{\contained_{R_{k-2}}(t)}$'' Sets.} Statement (2) of Lemma~\ref{lmm:hv:type2:contained-h*} allows us to produce $\contained_{R_{k-2}}(\bm{t})$ --- defined in \eqref{eqn:type2:contained-t} --- for each $\bm{t} \in \J(R_1, ..., R_{k-3}, H', V)$ as follows. First, fetch the (already computed) minimal segment $h^*$ of $\bm{t}[k-2]$ in $O(1)$ time. Then, scan the rectangles $r$ of $\contained_{R_{k-2}}(h^*)$ in descending order of $\xright(r)$. For each $r$ scanned, check whether $r \in \contained_{R_{k-2}}(\bm{t})$, or equivalently, whether $r$ covers $B_\bm{t}$ (recall that $B_\bm{t}$ is a point); the cost of this inspection is $O(1)$. Abort the scan as soon as $r \notin
\contained_{R_{k-2}}(\bm{t})$. This way, $\contained_{R_{k-2}}(t)$ can be decided in $O(k + |\contained_{R_{k-2}}(\bm{t})|)$ time. Doing so for all $\bm{t} \in \J(R_1, ..., R_{k-3},$ $H', V)$ takes $O(k \cdot |\J| + \sum_\bm{t} |\contained_{R_{k-2}}(\bm{t})|) = O(k \cdot \out)$ time, where the derivation used Lemma~\ref{lmm:hv:type2:recur-output} and statement (3) of Lemma~\ref{lmm:hv:type2:properties}.

@@

Let $P'$ be the set of remaining points in $P$ after the aforementioned removal. Next, for each $j \in [k-1]$, solve the reporting version of Problem $\mathscr{A}$ by feeding $P'$ and $R_j$ as the input. This produces the set $\contained_{R_j}(p)$ for each point $p \in P'$, where $\contained_{R_j}(p)$ is defined in \eqref{eqn:contained} (treating $p$ as a degenerated ``horizontal segment'') and includes all rectangles of $R_j$ covering $p$. By the discussion in Section~\ref{sec:bricks}, the total cost of this step is bounded by
\myeqn{
    O\Big(kn \log n + \sum_{p \in P'} \sum_{j \in [k-1]} |\contained_{R_j}(p)| \Big). \label{eqn:ksj:cat1:help1}
}

We are ready to generate $\J^\cat_k$. Take any point $p \in P'$, and let $r \in R_k$ be the rectangle with $p$ as the top-left corner. For every $(k-1)$-tuple
\myeqn{
    (r_1, ..., r_{k-1}) \in \contained_{R_1}(p) \times ... \times \contained_{R_{k-1}}(p) \nn
    %\label{eqn:ksj:cat1:help2}
}
we add $(r_1, ..., r_{k-1}, r)$ to $\J^\cat_k$. Performing the above for all $p \in P'$ generates the whole $\J^\cat_k$ in $O(1 + k \cdot|\J^\cat_k|)$ time. The way $P'$ is computed ensures that $\contained_{R_j}(p) \ne \emptyset$ for each $j \in [k-1]$. Hence, $\sum_{j \in [k-1]} |\contained_{R_j}(p)| \le \prod_{j \in [k-1]} |\contained_{R_j}(p)|$, which implies that \eqref{eqn:ksj:cat1:help1} is bounded by $O(kn \log n + k \cdot |\J^\cat_k|) = O(kn \log n + k \cdot \out)$ .

@@

Each rectangle of $R$ necessitates one insertion and one deletion in $\T$. All these insertions and deletions take $O(n \log n)$ time in total. Each segment $h$ of $H$ performs a reverse-containment query on $\T$, which has a cost of $O(\log n + K_h)$. All these queries demand a total cost of $O(n \log n + \sum_h K_h)$. The sum $\sum_h K_h$ here is equal to the total size of $\contained_R(h)$ for all the $h \in H$. The total size is equivalent to $\out$. The cost so far is therefore $O(n \log n + \out)$.

@@

Finally, we generate, for each $h$, its set $\contained_R(h)$ (i.e., the set of rectangles covering $h$), where the rectangles $r$ are sorted by $\xright(r)$. To start with, initialize an empty linked list for every $h \in H$. Inspect the pairs $(h, \lambda) \in L$ in ascending order of the ID-field $\lambda$. For each pair $(h, \lambda)$ examined, identify the rectangle $r$ whose ID is $\lambda$ and add $r$ to the linked list of $h$. By the way the rectangle IDs were generated, it is clear that $\xright(r)$ is larger than or equal to the x-coordinates of the right edges of the rectangles already in the linked list. The whole scan over $L$ finishes in $O(1 + \out)$ time, and produces the correct output for Problem $\mathscr{E}$.

@@

\begin{proposition} \label{prop:app:hv:type2:help2}
    For each tuple $\bm{t} \in \J(R_1,...,R_{k-3},H',V)$ (where $H'$ is defined in \eqref{eqn:hv:type2:H'}), the set $\contained_{R_{k-2}}(\bm{t})$ is non-empty.
\end{proposition}

\begin{proof}
    Let $h$ be the full segment of $\bm{t}[k-2]$. By the definition of a trimmed segment, there exists a rectangle $r_{k-2} \in R_{k-2}$ covering both endpoints of $h$. Thus, the effective horizontal segment of $\bm{t}$ (defined in Section~\ref{sec:hv:type2}), which is contained in $h'$, must also be covered by $r_{k-2}$. Hence, $\contained_{R_{k-2}}(\bm{t})$ contains $r_{k-2}$ and thus cannot be empty.
\end{proof}

@@

Consider any rectangle $r_{k-2} \in \contained_{R_{k-2}}(\bm{t})$. From the definition of $\contained_{R_{k-2}}(\bm{t})$, we know that $r_{k-2}$ covers the effective horizontal segment of $\bm{t}$ whose right endpoint is $h' \cap v$. This means that $h \cap v$ falls in $r_{k-2}$. We now have $h \, \cap \, v \, \cap \, \bigcap_{i = 1}^{k-2} r_i$ equals $h' \cap v$ and hence is non-empty, meaning that $(r_1,...,r_{k-3}, r_{k-2}, h, v) \in \J(R_1,...,R_{k-2},H,V)$.

@@

Next, we argue that $r_{k-2}\in \contained_{R_{k-2}}(\bm{t})$. Let $p'$ be the left endpoint of $h$. Because$h' \cap v = h \cap v = p$, the effective horizontal segment of $\bm{t}$ is the segment $p'p$. To prove $r_{k-2} \in \contained_{R_{k-2}}(\bm{t})$, we need to explain why $r_{k-2}$ covers this segment. This is true because:

@@

Next, assuming that the rectangles $r \in \contained_{R_{k-2}}(h^*)$ have been sorted in descending order of $\xright(r)$, we will prove that $\contained_{R_{k-2}}(\bm{t})$ includes a prefix of the sorted order. It suffices to establish the following equivalent claim:

>>>>

\def\xleft{\mathrm{left}}
\def\xright{\mathrm{right}}
\def\ybot{\mathrm{bot}}
\def\ytop{\mathrm{top}}
\def\maxtop{\mathrm{maxtop}}
\def\minleft{\mathrm{minleft}}
\def\gleft{\mathrm{left\text{-}guard}}
\def\gbot{\mathrm{bot\text{-}guard}}
\def\cross{\mathrm{cross}}
\def\dcross{\mathrm{d\text{-}cross}}
\def\contain{\mathrm{contain}}
\def\cat{\mathrm{cat1}}
\def\catt{\mathrm{cat2}}
\def\out{\mathrm{OUT}}
\def\side{\mathrm{side}}

@@

Given a horizontal segment $h = [x_1, x_2] \times y$, we say that a rectangle $r$ is a {\em left-end covering rectangle} of $h$ if $r$ contains the left endpoint of $h$ (i.e., $(x_1, y) \in r$). A horizontal/vertical segment $s$ {\em crosses} a rectangle $r$ if $s \cap r \ne \emptyset$ but $r$ covers neither of the two endpoints of $s$. A rectangle $r$ {\em contains} a horizontal/vertical segment $s$ if $r$ covers both endpoints of $s$.

@@

Let $S$ be a set of segments where either all segments are horizontal or all are vertical. Given a rectangle $r$, we define 
\myeqn{
    \cross_S(r) = \{s \in S \mid \text{$s$ crosses $r$} \};
    \label{eqn:cross}
}
namely, $\cross_S(r)$ is the set of segments in $S$ crossing $r$.
Let $R$ be a set of rectangles. Given a horizontal segment $h$, we define 
\myeqn{
    \contain_R(h) = \{r \in R \mid \text{$r$ contains $h$} \};
    \label{eqn:contained}
}
namely, $\contain_R(h)$ is the set of rectangles in $R$ containing $h$.

@@

\extraspacing {\bf Problem $\bm{\mathscr{E}}$.} The input involves a set $H$ of horizontal segments and a set $R$ of rectangles. The goal is to report, for each segment $h \in H$, the set $\contain_R(h)$ --- defined in \eqref{eqn:contained} --- where the rectangles are sorted by their right boundaries. Formally, if $r_1, r_2, ..., r_z$ for some $z \ge 1$ are all the rectangles in $\contain_R(h)$, we output $(h: r_1, r_2, ..., r_z)$, provided that $\xright(r_i) \ge \xright(r_{i-1})$ for each $i \in [2, z]$. Figure~\ref{fig:probs}e gives an example where $H = \set{h_1, h_2, h_3}$ and $R = \set{r_1, r_2, r_3}$; the output is $\set{(h_1: r_2, r_1), (h_2: r_2, r_3), (h_3: \emptyset)}$. In Appendix~\ref{app:bricks}, we explain how to solve the problem in $O(n \log n + \out)$ time where $n = |H| + |R|$ and $\out$ is the number of pairs $(h, r) \in H \times R$ such that $r$ contains $h$.

@@

\extraspacing {\bf Generating $\bm{\J_2}$.} Take any $(k-1)$-tuple $\bm{t} = (r_1, ..., r_{k-3}, h', v) \in \J(R_1, ..., R_{k-3}, H', V)$. Note that $B_\bm{t}$ --- defined in \eqref{eqn:B_t} --- is the point $h' \cap v$ (the intersection of $h'$ and $v$). Suppose $h' = [x_1, x_2] \times y$ and $B_\bm{t} = (x, y)$, we define the {\em effective horizontal segment} of $\bm{t}$ as the horizontal segment $[x_1, x] \times y$. This allows us to define
\myeqn{
    \contain_{R_{k-2}}(\bm{t}) = \{r \in R_{k-2} \mid \text{$r$ contains the effective horizontal segment of $\bm{t}$}   \}
    \label{eqn:type2:contained-t}
}
The above should not be confused with \eqref{eqn:contained}, where the ``$\contain$'' function takes a segment as the parameter, rather than a tuple.

\extraspacing {\underline{\em Example 5.3.}} Consider the tuples $\bm{t}_1$, $\bm{t}_2$, and $\bm{t}_3$ of $ \J(R_1, H', V)$ given in Example 5.2. For $\bm{t}_1 = (\alpha, h_2', v_1)$, its effective horizontal segment is $\texttt{DC}$ (see Figure~\ref{fig:hv:type2:ex}b). For $\bm{t}_2 = (\alpha, h_3', v_1)$, its effective horizontal segment is \texttt{EB}. For $\bm{t}_3 = (\alpha, h_3', v_2) \in \J(R_1, H', V)$, its effective horizontal segment is \texttt{EA}. Accordingly, as can be seen from Figure~\ref{fig:hv:type2:ex}c,
$\contain_{R_{k-2}}(\bm{t}_1) = \contain_{R_2}(\bm{t}_1) = \set{\beta_1}$, $\contain_{R_2}(\bm{t}_2) = \set{\beta_1, \beta_2, \beta_3}$, and $\contain_{R_2}(\bm{t}_3) = \set{\beta_3}$.
\done

@@

\begin{lemma} \label{lmm:hv:type2:properties}
    All the following statements are true:
    \myenums{
        \item Consider any $(k-1)$-tuple $\bm{t} \in \J(R_1, ..., R_{k-3}, H', V)$. Denote by $h$ the full segment of $\bm{t}[k-2]$. Then, for any $r \in \contain_{R_{k-2}}(\bm{t})$, the $k$-tuple $(\bm{t}[1], ..., \bm{t}[k-3], r, h, \bm{t}[k-1])$ must belong to $\J_2$.

        \item Consider any $k$-tuple $(r_1, ..., r_{k-2}, h, v) \in \J_2$. Let $h'$ be the trimmed segment of $h$ and set $\bm{t} = (r_1, ..., r_{k-3}, h', v)$. Then, $\bm{t} \in \J(R_1, ..., R_{k-3}, H', V)$ and $r_{k-2} \in \contain_{R_{k-2}}(\bm{t})$.

        \item $\sum_{\bm{t}} |\contain_{R_{k-2}}(\bm{t})| \le \out$, where the summation is over all $\bm{t} \in \J(R_1, ..., R_{k-3}, H', V)$.
    }
\end{lemma}

@@

\mytab{
    \> {\bf algorithm generate-$\J_2$} \\
    \> 1.\> $\J_2 = \emptyset$ \\
    \> 2.\> {\bf for} each $(k-2)$-tuple $\bm{t} \in \J(R_1, ..., R_{k-3}, H', V)$ {\bf do} \\
    \> 3.\>\> $h \leftarrow$ the full segment of $\bm{t}[k-2]$ \\
    \> 4.\>\> {\bf for} each $r \in \contain_{R_{k-2}}(\bm{t})$ {\bf do} \\
    \> 5.\>\>\> add $(\bm{t}[1], ..., \bm{t}[k-3], r, h, \bm{t}[k-1])$ to $\J_2$
}

@@

The correctness of the algorithm follows from statements (1) and (2) of Lemma~\ref{lmm:hv:type2:properties}. Furthermore, if we are given $\contain_{R_{k-2}}(\bm{t})$ for each $\bm{t}$, statement (3) of Lemma~\ref{lmm:hv:type2:properties}
assures us that the algorithm runs in $O(1 + |\J(R_1, ..., R_{k-3}, H', V)| + k \sum_\bm{t} |\contain_{R_{k-2}}(\bm{t})|) = O(1 + k \cdot \out)$ time, where the derivation used Lemma~\ref{lmm:hv:type2:recur-output} and statement (3) of Lemma~\ref{lmm:hv:type2:properties}.

\extraspacing {\underline{\em Example 5.4.}} For $\bm{t}_1 = (\alpha, h_2', v_1)$, the full segment of $h_2'$ is $h_2$. As $\beta_1$ is the only rectangle in $\contain_{R_2}(\bm{t}_1)$, Line 5 of the algorithm adds tuple $(\alpha, \beta_1, h_2, v_1)$ to $\J_2$. For $\bm{t}_2 = (\alpha, h_3', v_1)$, the full segment of $h_3'$ is $h_3$. As $\contain_{R_2}(\bm{t}_2) = \set{\beta_1, \beta_2, \beta_3}$, Line 5 adds $(\alpha, \beta_1, h_3, v_1)$, $(\alpha, \beta_2, h_3, v_1)$, and $(\alpha, \beta_3, h_3, v_1)$ to $\J_2$. Finally, the processing of $\bm{t}_3 = (\alpha, h_3', v_2)$ adds $(\alpha, \beta_3, h_3, v_2)$ to $\J_2$.
\done

@@

We are interested in the sets $\contain_{R_{k-2}}(h^*)$ of the segments $h^*$ in $H^*$, where $\contain_{R_{k-2}}(h^*)$ --- defined in \eqref{eqn:contained} --- is the set of rectangles in $R_{k-2}$ containing $h^*$. These sets have some useful properties:

\begin{lemma} \label{lmm:hv:type2:contained-h*}
    Both statements below are true: 
    \myenums{
        \item $\sum_{h^* \in H^*} |\contain_{R_{k-2}}(h^*)| \le \out$.

        \item Consider any tuple $\bm{t} \in \J(R_1, ..., R_{k-3}, H', V)$. Set $h' = \bm{t}[k-2]$ and let $h^*$ be the minimal segment of $h'$. Then, $$\contain_{R_{k-2}}(\bm{t}) \subseteq \contain_{R_{k-2}}(h^*).$$ Furthermore, if the rectangles $r$ in $\contain_{R_{k-2}}(h^*)$ are sorted in descending order of $\xright(r)$, then $\contain_{R_{k-2}}(\bm{t})$ includes a prefix of the sorted order.
    }
\end{lemma}

@@

\extraspacing {\underline{\em Example 5.6.}} It is clear from Figure~\ref{fig:hv:type2:ex}(d) that $\contain_{R_2}(h_3^*)$ has rectangles $\beta_3, \beta_2$, $\beta_1$, sorted in descending order of their right boundaries' x-coordinates. Consider $\bm{t}_2 = (\alpha, h_3', v_1)$ and $\bm{t}_3 = (\alpha, h_3', v_2)$. Segment $h_3^*$ is the minimal segment of $h_3'$. As stated in Lemma~\ref{lmm:hv:type2:contained-h*}, both $\contain_{R_2}(\bm{t}_2) = \set{\beta_1, \beta_2, \beta_3}$ and $\contain_{R_2}(\bm{t}_3) = \set{\beta_1}$ are prefixes of the sorted order of $\contain_{R_2}(h_3^*)$. Regarding $h_2^*$, it is the minimal segment of $h_2'$, and $\contain_{R_2}(h_2^*)$ contains only  $\beta_1$. For $\bm{t}_1 = (\alpha, h_2', v_1)$, $\contain_{R_2}(\bm{t}_1) = \set{\beta_1}$ is a (trivial) prefix of $\contain_{R_2}(h_2^*)$, as is also consistent with the lemma.
\done

Finding the $\contain_{R_{k-2}}(h^*)$ sets of all $h^* \in H^*$ is an instance of the find-all-sorted version of Problem $\mathscr{E}$ (with $H^*$ and $R_{k-2}$ as the input). The cost is $O(n \log n + \out)$ according statement (1) of Lemma~\ref{lmm:hv:type1:cross-r*} and the discussion in Section~\ref{sec:bricks}. Note that, for each $\contain_{R_{k-2}}(h^*)$ computed, the rectangles $r$ therein have been sorted in descending order of $\xright(r)$.

\extraspacing {\bf Computing the ``$\bm{\contain_{R_{k-2}}(t)}$'' Sets.} Statement (2) of Lemma~\ref{lmm:hv:type2:contained-h*} allows us to produce $\contain_{R_{k-2}}(\bm{t})$ --- defined in \eqref{eqn:type2:contained-t} --- for each $\bm{t} \in \J(R_1, ..., R_{k-3}, H', V)$ as follows. First, fetch the (already computed) minimal segment $h^*$ of $\bm{t}[k-2]$ in $O(1)$ time. Then, scan the rectangles $r$ of $\contain_{R_{k-2}}(h^*)$ in descending order of $\xright(r)$. For each $r$ scanned, check whether $r \in \contain_{R_{k-2}}(\bm{t})$, or equivalently, whether $r$ covers $B_\bm{t}$ (recall that $B_\bm{t}$ is a point); the cost of this inspection is $O(1)$. Abort the scan as soon as $r \notin
\contain_{R_{k-2}}(\bm{t})$. This way, $\contain_{R_{k-2}}(t)$ can be decided in $O(k + |\contain_{R_{k-2}}(\bm{t})|)$ time. Doing so for all $\bm{t} \in \J(R_1, ..., R_{k-3},$ $H', V)$ takes $O(k \cdot |\J| + \sum_\bm{t} |\contain_{R_{k-2}}(\bm{t})|) = O(k \cdot \out)$ time, where the derivation used Lemma~\ref{lmm:hv:type2:recur-output} and statement (3) of Lemma~\ref{lmm:hv:type2:properties}.

@@

Let $P'$ be the set of remaining points in $P$ after the aforementioned removal. Next, for each $j \in [k-1]$, solve the reporting version of Problem $\mathscr{A}$ by feeding $P'$ and $R_j$ as the input. This produces the set $\contain_{R_j}(p)$ for each point $p \in P'$, where $\contain_{R_j}(p)$ is defined in \eqref{eqn:contained} (treating $p$ as a degenerated ``horizontal segment'') and includes all rectangles of $R_j$ covering $p$. By the discussion in Section~\ref{sec:bricks}, the total cost of this step is bounded by
\myeqn{
    O\Big(kn \log n + \sum_{p \in P'} \sum_{j \in [k-1]} |\contain_{R_j}(p)| \Big). \label{eqn:ksj:cat1:help1}
}

We are ready to generate $\J^\cat_k$. Take any point $p \in P'$, and let $r \in R_k$ be the rectangle with $p$ as the top-left corner. For every $(k-1)$-tuple
\myeqn{
    (r_1, ..., r_{k-1}) \in \contain_{R_1}(p) \times ... \times \contain_{R_{k-1}}(p) \nn
    %\label{eqn:ksj:cat1:help2}
}
we add $(r_1, ..., r_{k-1}, r)$ to $\J^\cat_k$. Performing the above for all $p \in P'$ generates the whole $\J^\cat_k$ in $O(1 + k \cdot|\J^\cat_k|)$ time. The way $P'$ is computed ensures that $\contain_{R_j}(p) \ne \emptyset$ for each $j \in [k-1]$. Hence, $\sum_{j \in [k-1]} |\contain_{R_j}(p)| \le \prod_{j \in [k-1]} |\contain_{R_j}(p)|$, which implies that \eqref{eqn:ksj:cat1:help1} is bounded by $O(kn \log n + k \cdot |\J^\cat_k|) = O(kn \log n + k \cdot \out)$ .

@@

Each rectangle of $R$ necessitates one insertion and one deletion in $\T$. All these insertions and deletions take $O(n \log n)$ time in total. Each segment $h$ of $H$ performs a reverse-containment query on $\T$, which has a cost of $O(\log n + K_h)$. All these queries demand a total cost of $O(n \log n + \sum_h K_h)$. The sum $\sum_h K_h$ here is equal to the total size of $\contain_R(h)$ for all the $h \in H$. The total size is equivalent to $\out$. The cost so far is therefore $O(n \log n + \out)$.

@@

Finally, we generate, for each $h$, its set $\contain_R(h)$ (i.e., the set of rectangles covering $h$), where the rectangles $r$ are sorted by $\xright(r)$. To start with, initialize an empty linked list for every $h \in H$. Inspect the pairs $(h, \lambda) \in L$ in ascending order of the ID-field $\lambda$. For each pair $(h, \lambda)$ examined, identify the rectangle $r$ whose ID is $\lambda$ and add $r$ to the linked list of $h$. By the way the rectangle IDs were generated, it is clear that $\xright(r)$ is larger than or equal to the x-coordinates of the right edges of the rectangles already in the linked list. The whole scan over $L$ finishes in $O(1 + \out)$ time, and produces the correct output for Problem $\mathscr{E}$.

@@

\begin{proposition} \label{prop:app:hv:type2:help2}
    For each tuple $\bm{t} \in \J(R_1,...,R_{k-3},H',V)$ (where $H'$ is defined in \eqref{eqn:hv:type2:H'}), the set $\contain_{R_{k-2}}(\bm{t})$ is non-empty.
\end{proposition}

\begin{proof}
    Let $h$ be the full segment of $\bm{t}[k-2]$. By the definition of a trimmed segment, there exists a rectangle $r_{k-2} \in R_{k-2}$ covering both endpoints of $h$. Thus, the effective horizontal segment of $\bm{t}$ (defined in Section~\ref{sec:hv:type2}), which is contained in $h'$, must also be covered by $r_{k-2}$. Hence, $\contain_{R_{k-2}}(\bm{t})$ contains $r_{k-2}$ and thus cannot be empty.
\end{proof}

@@

Consider any rectangle $r_{k-2} \in \contain_{R_{k-2}}(\bm{t})$. From the definition of $\contain_{R_{k-2}}(\bm{t})$, we know that $r_{k-2}$ covers the effective horizontal segment of $\bm{t}$ whose right endpoint is $h' \cap v$. This means that $h \cap v$ falls in $r_{k-2}$. We now have $h \, \cap \, v \, \cap \, \bigcap_{i = 1}^{k-2} r_i$ equals $h' \cap v$ and hence is non-empty, meaning that $(r_1,...,r_{k-3}, r_{k-2}, h, v) \in \J(R_1,...,R_{k-2},H,V)$.

@@

Next, we argue that $r_{k-2}\in \contain_{R_{k-2}}(\bm{t})$. Let $p'$ be the left endpoint of $h$. Because$h' \cap v = h \cap v = p$, the effective horizontal segment of $\bm{t}$ is the segment $p'p$. To prove $r_{k-2} \in \contain_{R_{k-2}}(\bm{t})$, we need to explain why $r_{k-2}$ covers this segment. This is true because:

@@

Next, assuming that the rectangles $r \in \contain_{R_{k-2}}(h^*)$ have been sorted in descending order of $\xright(r)$, we will prove that $\contain_{R_{k-2}}(\bm{t})$ includes a prefix of the sorted order. It suffices to establish the following equivalent claim:
